Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-06 ms.tex
6.3

It is also possible that many of the \citet{Miller+15} sample of
galactic nuclei are not accreting in steady state.  The value of the
X-ray efficiency assumed in the right panel of Figure \ref{fig:miller}
represents a maximum value, since it does not account for suppression
of the SMBH accretion rate due to outflows from the accretion disk on
small radial scales (e.g., \citealt{Blandford&Begelman99}).  If the
actual radiative efficiency is lower than our fiducial assumption,
then many of the X-ray detections would lie above the predicted value
of $\langle L_X \rangle$ for stable accretion (black line).  In this
case the X-ray detected galaxies either experience less heating than
average galaxies of the same mass, or they are not accreting stably at
all.  {\bf BDM: we may need to remove - why did this change so much?}
Indeed, for $M_{\bullet} \lesssim 10^{8}M_{\odot}$, $L_{\rm X}$ nearly
exceeds that predicted for the maximum thermally-stable accretion
rate, $\dot{M}_{\rm TI}$ (yellow line).  Since the \citet{Miller+15}
sample includes only early-type galaxies with older stellar
populations than the average for their black hole mass, it would not
be too surprising if many of these galaxies were thermally unstable.

Figure \ref{fig:bh_growth} (top panel) also shows that $\dot{M}_{\rm
  C} > \dot{M}_{\rm TI}$ across all black hole masses.  Thus, even if
Compton heating could act to stabilize the flow (which, as already
mentioned, is unlikely), for fiducial parameters it is insufficient to
produce a thermally stable accretion rate given conditions
corresponding to the average star formation history.  When thermal
instabilities become manifest, nuclear gas may undergo cyclic
oscillations of rapid cooling and high accretion rates, followed by
quiescent periods once gas has been consumed and/or AGN feedback
becomes effective.  Such cycles are seen in numerical simulations of
elliptical galaxies on large scales and long timescales
(\citealt{Ciotti+10}), an idea which is supported by observations
showing that a fraction of early-type galaxies in the local universe
have undergone recent ($< 1-2$ Gyr old) star formation episodes
(\citealt{Donas+07}).

Similar cyclic AGN activity could occur on the smaller radial scale of
the sphere of influence (e.g.~\citealt{Yuan&Li11},
\citealt{Cuadra+15}).  In this case the large scatter in the X-ray
luminosities shown in Fig \ref{fig:miller} could result from the wide
range of inflow rates experienced over the course of a cyclic episode
between instabilities and periodic nuclear outbursts.  Due to their
deeper potential well on scales of the sphere of influence, core galaxies are more likely to possess higher accretion rates and
potentially be thermally unstable for a fixed wind heating rate than
cusp galaxies.

@@

If the galaxies for which $L_{\rm X}$ is measured at low $M_{\bullet}$ in Fig
\ref{fig:miller} are indeed in a state of thermally-unstable cyclic behavior, then a subset of the galaxies with X-ray upper limits may contribute to a separate steady-state, thermally stable population, for instance including SgrA* (which has a much lower X-ray luminosity for its black hole mass than predicted by the \citet{Miller+15} trend).  One consequence of a potentially bimodal population of steady and cyclic galactic nuclei calls into question the physical justification of using simple extrapolations of power-law fits to the $L_{\rm X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to constrain the occupation fraction of SMBHs in the nuclei of low mass galaxies.

@@

The bottom panel of Figure~\ref{fig:bh_growth} shows the SMBH growth
time, $t_{\rm grow} = M_{\bullet}/\dot{M}$, for each accretion rate
shown in the top panel.  For black holes accreting steadily given the
stellar wind heating associated with the average star formation
history of their galaxies, we see that $t_{\rm grow}$ exceeds the
Hubble time by 1$-$3 orders of magnitude across all black hole masses.
Steady accretion therefore cannot explain the growth of low mass black
holes, a fact which is not surprising given that approximately half of
this growth occurs in AGN radiating within 10 per cent of their
Eddington rate (\citealt{Heckman+04}).  Such high accretion rates
likely instead require an source of gas external to the nuclear
region, triggered either by galaxy mergers associated with the
hierarchical growth of structure or thermal instabilities on larger,
galactic scales (e.g.~\citealt{Ciotti+10}).  In support of this,
\citet{Kauffmann&Heckman09} find that the growth of low mass black
holes is dominated by young stellar bulges which have a log normal
distribution of Eddington ratios peaked at $\Mdot/\MdotEdd\sim
0.01$. That this distribution is insensitive to galaxy properties
suggests that it is set by SMBH feedback instead of the nuclear gas
supply.

>>>>

It is also possible that many of the \citet{Miller+15} sample of
galactic nuclei are not accreting in steady state.  The value of the
X-ray efficiency assumed in the right panel of Figure \ref{fig:miller}
represents a maximum value, which does not for instance account for suppression
of the SMBH accretion rate due to outflows from the accretion disk on
small radial scales (e.g., \citealt{Blandford&Begelman99}).  Furthermore, our SFH-averaged accretion rates are derived using analytic expressions for the stagnation radius that may underpredict the true accretion rate if $\zeta < \zeta_c$, as may occur in massive core galaxies with large break radii.  If the actual radiative efficiency is lower than our fiducial assumption,
then many of the X-ray detections would lie above the predicted value
of $\langle L_X \rangle$ for stable accretion (black line).  In this
case the X-ray detected galaxies experience less heating than
average galaxies of the same mass, and may not be stably accreting at
all.  This is particularly plausible because the \citet{Miller+15} sample includes only early-type galaxies, with older stellar
populations than the average for their black hole mass.

Figure \ref{fig:bh_growth} (top panel) also shows that $\dot{M}_{\rm
  C} > \dot{M}_{\rm TI}$ across all black hole masses.  Thus, even if
Compton heating could act to self-stabilize the flow (which, as already
mentioned, appears unlikely), its magnitude appears to be insufficient to
produce a thermally stable accretion rate even given conditions
corresponding to the average star formation history.

If prsent, thermal
instabilities may cause the CNM to undergo cyclic
oscillations of rapid cooling and high accretion rates, followed by
quiescent periods once gas has been consumed and/or AGN feedback
becomes effective.  Such cycles are seen in numerical simulations of
elliptical galaxies on large scales and long timescales
(\citealt{Ciotti+10}).  Evidence for such periodic outbursts includes observations
showing that a fraction of early-type galaxies in the local universe
have undergone recent ($< 1-2$ Gyr old) star formation episodes
(\citealt{Donas+07}).

Similar cyclic AGN activity could occur on the smaller radial scale of
the sphere of influence (e.g.~\citealt{Yuan&Li11},
\citealt{Cuadra+15}).  In this case the large scatter in the X-ray
luminosities shown in Fig \ref{fig:miller} could result from the wide
range of inflow rates experienced over the course of a cyclic episode
between instabilities and periodic nuclear outbursts.

% Due to their
%deeper potential well on scales of the sphere of influence, core galaxies are more likely to possess higher accretion rates and
%potentially be thermally unstable for a fixed wind heating rate than
%cusp galaxies.

@@

If the galaxies for which $L_{\rm X}$ is measured at low $M_{\bullet}$ in Fig
\ref{fig:miller} are indeed in a state of thermally-unstable cyclic behavior, then a subset of the galaxies with X-ray upper limits may contribute to a separate population of thermally stable, steadily accreting nuclei.  Such a population could include SgrA*, which has a much lower X-ray luminosity for its black hole mass than predicted by the \citet{Miller+15} trend.  A potentially bimodal population of steady and cyclic galactic nuclei calls into question the physical justification of using simple extrapolations of power-law fits to the $L_{\rm X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to constrain the occupation fraction of SMBHs in the nuclei of low mass galaxies.

@@

The bottom panel of Figure~\ref{fig:bh_growth} shows the SMBH growth
time, $t_{\rm grow} = M_{\bullet}/\dot{M}$, for each accretion rate
shown in the top panel.  For black holes accreting steadily at the rate set by the
stellar wind heating associated with the average star formation
history of their host galaxies, we see that $t_{\rm grow}$ exceeds the
Hubble time by 2$-$3 orders of magnitude across all black hole masses.
Steady accretion therefore cannot explain the growth of low mass black
holes, a fact which is not surprising given that approximately half of
this growth occurs in AGN radiating within 10 per cent of their
Eddington rate (\citealt{Heckman+04}).  Such high accretion rates
likely instead require an source of gas external to the nuclear
region, triggered either by galaxy mergers associated with the
hierarchical growth of structure or thermal instabilities on larger,
galactic scales (e.g.~\citealt{Ciotti+10}).  In support of this,
\citet{Kauffmann&Heckman09} find that the growth of low mass black
holes is dominated by young stellar bulges which have a log normal
distribution of Eddington ratios peaked at $\Mdot/\MdotEdd\sim
0.01$. That this distribution is insensitive to galaxy properties
suggests that it is set by SMBH feedback instead of the nuclear gas
supply.
